Efficient pedigree recording for fast population genetics simulation

### 2017-11-22 forwards_paper.tex
Updated Algorithm W to include simplify.
Also general editing pass.

\include{macros}
\newcommand{\simupop}{\texttt{simuPOP}}
\newcommand{\fwdpp}{\texttt{fwdpp}}
\newcommand{\msprime}{\texttt{msprime}}
\newcommand{\nodes}{\texttt{nodes}}
\newcommand{\edgesets}{\texttt{edgesets}}
\newcommand{\sites}{\texttt{sites}}
\newcommand{\mutations}{\texttt{mutations}}
\newcommand{\setdiff}{\smallsetminus}
\newcommand{\Nt}{\mathcal{N}}  % node table
\newcommand{\Et}{\mathcal{E}}  % edge table
\newcommand{\St}{\mathcal{S}}  % site table
\newcommand{\Mt}{\mathcal{M}}  % mutation table
\newcommand{\Al}{\mathcal{A}}  % list of ancestors
\DeclareMathOperator{\len}{length}
\DeclareMathOperator{\add}{add} % for adding to tables
\newcommand{\lleft}{\mathrm{left}}  % left label
\newcommand{\lright}{\mathrm{right}}
\newcommand{\lid}{\mathrm{id}}

@@

To record the genealogical history of a forwards in time simulation
we need to record two things for each new chromosome:
the birth time, and the endpoints and parental IDs of each distinctly inherited segment,
which are naturally stored as the \emph{nodes} and \emph{edges} of a tree sequence.
For concreteness, here we write out in pseudocode how to run a neutral Wright--Fisher simulation
(but with overlapping generations) that records genealogical history in this way.
The simulation will run for $T$ generations,
and has $N$ haploid individuals, each carrying a single chromosome of length L.
For simplicity we sample exactly one crossover per generation.
The probability of death per individual each generation is $\delta$.
We will use
$\randomuniform(A)$ to denote an element of the set $A$ chosen uniformly at random
(and each such instance represents and independent draw).

\jda{TODO: rewrite with tables}
We begin in~\algref{W1} by allocating our initial population $P$
(a vector of $N$ node IDs),
and creating $N$ nodes with birth time $T$ generations ago, recorded in the
vector $\tau$. The set $E$ is used to store the edges that we output during the
simulation, and $n$ is the number of nodes created so far (and so, the ID of
the next node we create). Steps~\algref{W2} and~\algref{W3} simply loop over
the generation clock $t$ and individual index $j$. In~\algref{W4} we check if
an individual $P_j$ has died in this generation. If it has, we replace it in
steps \algref{W5}--\algref{W7}; if not, we proceed immediately to the next
individual. When an individual in the population dies, we first allocate
a new node with ID $n$ in~\algref{W5} and record its birth time. Then,
in step~\algref{W6} we choose two indexes $a$ and $b$ uniformly (giving us
parental IDs $P_a$ and $P_b$) and choose a breakpoint $x$. We record the effects of this
event by storing two new edges: one recording that the parent of node $n$
from $0$ to $x$ is $P_a$, and another recording that the parent of $n$
from $x$ to $L$ is $P_b$. We then complete the replacement event by incrementing
$n$, ready to represent the next new node.

\begin{taocpalg}{W}{Forwards-time tree sequence}
{Simulates a randomly mating population $P$ of $N$ haploid individuals,
with a probability of death per generation $\delta$ and a chromosome of length
$L$ for $T$ generations. Events are recorded in a node table $\mathcal{N}$
and edge table $\mathcal{E}$. Simplify is run every $s$ generations.
}

\algstep{W1.}{Initialisation.}{
Set $\mathcal{N} \leftarrow \mbox{NodeTable}()$ and $\mathcal{E}
\leftarrow \mbox{EdgeTable}()$ and $t \leftarrow T$.
 For $0 \leq j < N$, set $P_j \leftarrow \mbox{add}(\mathcal{N}, T)$.
}

\algstep{W2.}{Generation loop.}{ If $t = 0$ terminate.
Set $P' \leftarrow P$ and then set $t \leftarrow t - 1$ and $j \leftarrow 0$.
}

\algstep{W3.}{Individual loop.}{ If $j = N$ set $P \leftarrow P'$ and go to \algref{W2}.
}

\algstep{W4.}{Mortality.}{ If $\randomuniform([0, 1)) \geq \delta$ go to \algref{W8}.
}

\algstep{W5.}{New node.}{Set $u \leftarrow \mbox{add}(\mathcal{N}, t)$ and $P'_j \leftarrow u$.
}

\algstep{W6.}{Choose parents.}{Set $a \leftarrow \randomuniform(\{0, \dots, N - 1\})$,
    $b \leftarrow \randomuniform(\{0, \dots, N - 1\})$ and $x \leftarrow \randomuniform([0, L))$.
}

\algstep{W7.}{Record edges}{
Call $\mbox{add}(\mathcal{E}, (0, x, P_a, u))$ and $\mbox{add}(\mathcal{E}, (x,
L, P_b, u))$.
}

\algstep{W8.}{Loop tail}{Set $j \leftarrow j + 1$ and go to \algref{W3}.
}

\algstep{W9.}{Simplify}{TODO}

@@

This algorithm records only the topological information resulting from the
forwards in time Wright-Fisher process, but it is straightforward to add
mutational information. This can be done in two different ways.
We can record mutations that occur during the simulation quite simply.
For example, in Algorithm~\algref{W} we would generate mutations after
we have recorded the edges joining the new node $n$ to its parents
$P_a$ and $P_b$ in step \algref{W7}.
For example, if we assume that mutations occur according to the infinite sites model
at rate $\mu$ per generation per unit length then there are
$\mbox{Poisson}\left(\mu (r - \ell)(\tau_p
- \tau_c)\right)$ mutations on each edge $(\ell, r, p, c)$. Each of these mutations
will occur on a distinct site $x$ drawn uniformly from $(\ell, r]$ and be associated with node $c$.
It is straightforward to record this information during the simulation, but it
is significantly simpler and more efficient to generate these mutations
after the simulation has completed.
% In the case of infinite sites mutations
% there is no difference between generating these mutations during the forwards
% in time simulation and after it has completed. We are using precisely the same
% information.

@@

Besides the cost of transmitting genotypes,
generating mutations as a separate process after the topological simulation has completed
avoids generating the many mutations that are lost in the population.
Figure~\ref{fig:wf-trees} shows
an example of a marginal genealogy produced by a forwards-time Wright--Fisher
process like Algorithm~\algref{W}.
On the left is the tree showing all the edges output by the simulation,
while on the right
is the minimal tree representing the ancestry of the currently alive
population. Clearly there is a great deal of redundancy in the topological
information output by the simulation.

@@

A minimal genealogy such as in Figure~\ref{fig:wf-trees}
has fewer edges on which to place mutations.
There are many more advantages to simplifying the genealogies to this point.
Computing this minimal representation of the
edges output by a forward-time simulation is an instance of a more
general problem that we cover in the next section.

@@

Suppose that we are only interested in a subset of the nodes (our `samples')
of a tree sequence,
and so wish to reduce this input tree sequence
into the minimum representation of the topologies that include the specified
samples. The output tree sequence must have the following properties:
\begin{enumerate}
\item All marginal trees must match the subtree induced by the samples of the corresponding tree in the input tree sequence.
\item Within the marginal trees, all non-sample vertices must has at least
two children (i.e., unary tree vertices are removed), unless the vertex is a sample.
\item Any nodes and edges not ancestral to the sampled nodes are removed.
\item There are no adjacent redundant edges, i.e, pairs of edges $(\ell, x, p,
c)$ and $(x, r, p, c)$ which can be represented with a single edge
$(\ell, r, p, c)$.
\end{enumerate}

@@

More concretely,
the algorithm works by moving back through time,
processing each parent in the input tree sequence in chronological order.
The main state of the algorithm at each point in time is a set of ancestral lineages,
and each lineage is a collection of ancestral segments.
An ancestral segment $(\ell, r, u)$ is found in a lineage
if the output node $u$ inherits the genomic interval $[\ell, r)$ from that lineage.
% These segments are stored in a collection of linked lists,
% one list of segments for each ancestral lineage present at that time.
(These lineages are the ``colors'' above.)
We also maintain a map $A$ such that $A_j$ is segment chain for node $j$ in
the input tree sequence.
Crucially, the time required to run the algorithm is linear in the number of edges of the input tree sequence.

@@

\jda{It'd be good to outline what readers should expect here.}
\jda{and note we're doing the space analysis for with reference to alg W with $s=1$}
Consider a simulation of a Wright-Fisher population of $2N$ haploid individuals
using Algorithm~\algref{W}
for $T$ generations,
with the probability of death, $\delta$, set to 1 for simplicity.
Since there is exactly one crossing over every generation,
this produces tables of
$2NT$ nodes and
$4NT$ edges.
Suppose that $T$ is large enough that all samples coalesce within the simulation with high probability,
so that $T \sim 20N$, say.
After simplification, we are left with the tree sequence describing the history
of only the current generation of $2N$ individuals.
This tree sequence has $4N-2$ edges to describe the leftmost tree;
and each time the marginal tree changes along the sequence,
three edges end and three new edges begin (except for changes affecting the root; see \citet{kelleher2016efficient}).
Coalescent theory tells us that
the expected total length of edges in a marginal tree is approximately $4N\log(2N)$,
which is also equal to the mean number of ancestral recombination events that occur on a branch of the marginal tree,
since we have taken one crossover per generation.
Not all such recombinations actually change the tree topology,
and three times this gives us an upper bound on the expected number of edges.
Similarly, not every new edge derives from a never-before-seen node,
but the number of nodes is at most equal to the number of edges plus the sample size.
With $T=20N$, this reduces the initial storage of $160 N^2$ items to $8N(4\log(2N) + 3)$;
at $N=10^4$, a factor of 4,700.

To get an idea of how required space depends on the length of time,
suppose instead that we have run a simulation of $2N$ individuals for $T$ generations,
begun with no prior history.
This produces $4NT$ edges; how many are required after simplification?
As above, the expected number of edges is bounded by $2N-2$ plus four times the mean marginal tree length.
Roughly, the length of time for which a marginal tree has k tips is
$2N/k(k-1) = 2N(1/(k-1) - 1/k)$,
and so the time to go from N to n tips is $2N(1/(n-1) - 1/(N-1))$.
This implies that after running for time $T$ we expect to
have around $n(T)$ roots, where $n(T) \approx 1 + 2N/(T+2)$.
The total tree length over this time is
$2 N \sum_{k=n(T)+1}^{N-1} 1/k$, which
% is approximately
% \begin{align*}
%     2 N \log\left( \frac{N}{1 + 2 N/(T+2)} \right) .
% \end{align*}
leads to an upper bound on the number of edges of
\begin{align*}
    2 N \left( 1 + 4 \log\left( \frac{NT}{T + 2 N} \right)\right) .
\end{align*}
This holds up quite well in practice --
the number of edges actually required for 50 independent Wright--Fisher simulations,
as a function of time (obtained by running simplify every generation)
is shown in \ref{fig:num_edges}, compared to this prediction.

@@

In this paper, we have shown that storing pedigrees
and associated recombination events
in a forwards-time simulation
not only results in having available a great deal more information about the simulation,
but also can speed up the simulation by orders of magnitude.
To make this feasible,
we have described how to efficiently store this information in tables,
and have described a fundamental algorithm for simplificication of tree sequences.
Conceptually, recording of genealogical and recombination events
can happen independently of the details of simulation;
for this reason, we provide a well-defined and well-tested API in C and in python
for use in other code bases.

The tree sequences produced by default by this method
are very compact, storing genotype \emph{and} genealogical information
in a small fraction of the space taken by a compressed VCF file.
The format is also highly efficient to process,
leading to advantages for downstream analysis as well.
This is because many algorithms to compute statistics of interest for population genetics
are naturally expressed in terms of tree topologies,
and so can be quickly computed from the trees underlying the tree sequence format.
For example, pairwise nucleotide diversity $\pi$, is the average density of
differences between sequences in the sample.
To compute this directly from sequence data at $m$ sites in $n$ samples
requires computing allele frequencies, taking $O(nm)$ operations.
By using the locations of the mutations on the marginal trees,
and the fact that these are correlated,
dynamic programming algorithms similar to those in~\citep{kelleher2016efficient}
can do this in roughly $O(n + m \log n)$ operations.
The \msprime\ API provides a method to compute $\pi$ among arbitrary subsets of the
samples in a tree sequence, which took about 1.2 seconds
applied to the example tree sequence above with 1.1 million mutations
in 200 megabases for 500,000 samples.

@@

The Tables API is primarily designed to facilitate efficient interchange of
data between programs or between different modules of the same program.
\jda{should capitalize Tables, or not, as above}
Following the current best-practises [citations: apache arrow, etc] data is stored
in a columnar format.
The principal advantage of this for our purposes is that it allows for very efficient
interchange of large amounts of numerical data. In principle, this enables
zero-copy semantics, where a data consumer can read the information directly
from the memory of a producer without incurring the overhead of a copy
[citation?] Our implementation uses the numpy C API [citation] to efficiently copy
data from Python into the low-level C library used to manipulate
tree sequences.
Thus, by using a simple numerical
encoding of tree topologies and contiguous arrays of data to store this
encoding, we can achieve data transfer rates that would be impossible under
any text-based approach while retaining excellent portability.

@@

>>>>

\include{macros}
\newcommand{\simupop}{\texttt{simuPOP}}
\newcommand{\fwdpp}{\texttt{fwdpp}}
\newcommand{\msprime}{\texttt{msprime}}
\newcommand{\nodes}{\texttt{nodes}}
\newcommand{\edgesets}{\texttt{edgesets}}
\newcommand{\sites}{\texttt{sites}}
\newcommand{\mutations}{\texttt{mutations}}
\newcommand{\setdiff}{\smallsetminus}
\newcommand{\Nt}{\mathcal{N}}  % node table
\newcommand{\Et}{\mathcal{E}}  % edge table
\newcommand{\St}{\mathcal{S}}  % site table
\newcommand{\Mt}{\mathcal{M}}  % mutation table
\newcommand{\Al}{\mathcal{A}}  % list of ancestors
\newcommand{\taddnode}[1]{\mbox{addnode}(#1)}
\newcommand{\taddedge}[1]{\mbox{addedge}(#1)}
\newcommand{\tsimplify}[1]{\mbox{simplify}(#1)}
\DeclareMathOperator{\len}{length}
\DeclareMathOperator{\add}{add} % for adding to tables
\newcommand{\lleft}{\mathrm{left}}  % left label
\newcommand{\lright}{\mathrm{right}}
\newcommand{\lid}{\mathrm{id}}

@@

To record the genealogical history of a forwards in time simulation
we need to record two things for each new chromosome:
the birth time, and the endpoints and parental IDs of each distinctly inherited segment,
which are naturally stored as the \emph{nodes} and \emph{edges} of a tree sequence.
For concreteness, here we write out in pseudocode how to run a neutral Wright--Fisher simulation
(but with overlapping generations) that records genealogical history in this way.
The simulation will run for $T$ generations,
and has $N$ haploid individuals, each carrying a single chromosome of length L.
For simplicity we sample exactly one crossover per generation.
The probability of death per individual each generation is $\delta$.
We will use
$\randomuniform(A)$ to denote an element of the set $A$ chosen uniformly at random
(and each such instance represents and independent draw). Given a node table
$\Nt$, the function $\taddnode{\Nt, t}$ adds a new node to the table with time $t$
and returns the ID of this new node. Similarly, the function $\taddedge{\Et,
\ell, r, p, c}$ adds a new edge $(\ell, r, p, c)$ to the edge table $\Et$.
The function $\tsimplify{P, \Nt, \Et}$ simplifies the history stored in the
tables $\Nt$ and $\Et$ to the minimal information required to represent the
genealogies of the list of node IDs $P$ (see section XXX).
We first list
the algorithm in detail, and immediately follow this with a step-by-step
explanation.

\begin{taocpalg}{W}{Forwards-time tree sequence}
{Simulates a randomly mating population of $N$ haploid individuals with
chromosome of length $L$ for $T$ generations, and returns the node
and edge tables ($\Nt$ and $\Et$) recording the simulated history.
Simplify is run every $s$ generations, removing genealogical
information from $\Nt$ and $\Et$ unreachable from the current population $P$.
}

\algstep{W1.}{Initialisation.}{
Set $\Nt \leftarrow \mbox{NodeTable}()$, $\mathcal{E}
\leftarrow \mbox{EdgeTable}()$ and $t \leftarrow T$.
 For $0 \leq j < N$, set $P_j \leftarrow \taddnode{\Nt, T}$. Set $j \leftarrow 0$.
}

\algstep{W2.}{New node.}{Set $u \leftarrow \taddnode{\Nt, t}$ and $P'_j \leftarrow u$.
}

\algstep{W3.}{Choose parents.}{Set $a \leftarrow \randomuniform(\{0, \dots, N - 1\})$,
    $b \leftarrow \randomuniform(\{0, \dots, N - 1\})$ and $x \leftarrow \randomuniform([0, L))$.
}

\algstep{W4.}{Record edges}{
Call $\taddedge{\Et, 0, x, P_a, u}$ and $\taddedge{\Et, x, r, P_b, u}$.
}

\algstep{W5.}{Individual loop}{Set $j \leftarrow j + 1$. If $j < N$ go to \algref{W2}.
Otherwise, if $t\bmod s \neq 0$ go to \algref{W7}.
}

\algstep{W6.}{Simplify.}{Call $\tsimplify{P', \Nt, \Et}$, and set $P'_j
\leftarrow j $ for $0 \leq j < N$. (Tables may need to be sorted.)}

\algstep{W7.}{Generation loop.}{Set $t \leftarrow t - 1$. If $t = 0$ terminate.
Set $P \leftarrow P'$, $j \leftarrow 0$, and go to \algref{W2}.
}

@@

We begin in~\algref{W1} by creating new node and edge tables, and setting
our population $P$ (a vector of $N$ node IDs) to the initial population.
This initial population is a set of $N$ nodes with birth time $T$ generations
ago. We also initialise our generation clock $t$ and individual index $j$.
Steps~\algref{W2}--\algref{W4} replace the individual with node ID
$P_j$ by first creating a new node with birth time $t$ (and ID $u$).
Then, in step~\algref{W3} we choose two indexes $a$ and $b$ uniformly (giving us
parental IDs $P_a$ and $P_b$) and choose a breakpoint $x$. We record the effects of this
event by storing two new edges: one recording that the parent of node $u$
from $0$ to $x$ is $P_a$, and another recording that the parent of $u$
from $x$ to $L$ is $P_b$. Step \algref{W5} then iterates over each of the
$N$ individuals for each generation. At the end of a generation, we then check
if we need to simplify (which is done every $s$ generations). If simplification
is required, we do this in step \algref{W6} by calling the simplify function
on the node and edge tables with the current set of population IDs $P'$ as
the samples. This updates the tables in-place to remove all redundant
information, and remaps the specified sample IDs to $0, \dots N - 1$
in the updated tables. Hence, we set our current population IDs to
$0, \dots N - 1$ after simplify has completed. (See section XXX for more
details on simplify.) Step \algref{W7} then completes the algorithm by looping
over generations; we decrement our clock $t$, and terminate if $t = 0$.
Otherwise, we update our current population and individual index and loop
back to \algref{W2}.

@@

This algorithm records only the topological information resulting from the
forwards in time Wright-Fisher process, but it is straightforward to add
mutational information. This can be done in two different ways.
We can record mutations that occur during the simulation quite simply.
For example, in Algorithm~\algref{W} we would generate mutations after
we have recorded the edges joining the new node $n$ to its parents
$P_a$ and $P_b$ in step \algref{W7}.
It is straightforward to record this information during the simulation, but it
is significantly more efficient to generate these mutations
after the simulation has completed becuase it
avoids the cost of generating the many mutations that are lost in the population.
Figure~\ref{fig:wf-trees} shows
an example of a marginal genealogy produced by a forwards-time Wright--Fisher
process like Algorithm~\algref{W}.
On the left is the tree showing all the edges output by the simulation,
while on the right
is the minimal tree representing the ancestry of the currently alive
population. Clearly there is a great deal of redundancy in the topological
information output by the simulation.

@@

@@

Suppose that we are only interested in a subset of the nodes (our `samples')
of a tree sequence,
and so wish to reduce this input tree sequence
into the minimum representation of the topologies that include the specified
samples. The output tree sequence must have the following properties:
\begin{enumerate}
\item All marginal trees must match the subtree induced by the samples of the corresponding tree in the input tree sequence.
\item Within the marginal trees, all non-sample vertices must has at least
two children (i.e., unary tree vertices are removed).
\item Any nodes and edges not ancestral to the sampled nodes are removed.
\item There are no adjacent redundant edges, i.e, pairs of edges $(\ell, x, p,
c)$ and $(x, r, p, c)$ which can be represented with a single edge
$(\ell, r, p, c)$.
\end{enumerate}

@@

\jk{Insert lost text about simplify efficiency and plot}

@@

\jda{It'd be good to outline what readers should expect here.}
\jda{and note we're doing the space analysis for with reference to alg W with $s=1$}
Consider a simulation of a Wright-Fisher population of $2N$ haploid individuals
using Algorithm~\algref{W}
for $T$ generations and simplify interval $s = 1$ so that redundant information
is removed after every generation.
Since there is exactly one crossing over every generation,
this produces tables of
$2NT$ nodes and
$4NT$ edges.
Suppose that $T$ is large enough that all samples coalesce within the simulation with high probability,
so that $T \sim 20N$, say.
After simplification, we are left with the tree sequence describing the history
of only the current generation of $2N$ individuals.
This tree sequence has $4N-2$ edges to describe the leftmost tree;
and each time the marginal tree changes along the sequence,
four edges end and four new edges begin (except for changes affecting the root; see \citet{kelleher2016efficient}).
Coalescent theory tells us that
the expected total length of edges in a marginal tree is approximately $4N\log(2N)$,
which is also equal to the mean number of ancestral recombination events that occur on a branch of the marginal tree,
since we have taken one crossover per generation.
Not all such recombinations actually change the tree topology,
and four times this gives us an upper bound on the expected number of edges.
Similarly, not every new edge derives from a never-before-seen node,
but the number of nodes is at most equal to the number of edges plus the sample size.
With $T=20N$, this reduces the initial storage of $160 N^2$ items to $8N(4\log(2N) + 3)$;
at $N=10^4$, a factor of 4,700.
\jk{I haven't checked these numbers, just corrected the obvious threes/fours.}

To get an idea of how required space depends on the length of time,
suppose instead that we have run a simulation of $2N$ individuals for $T$ generations,
begun with no prior history.
This produces $4NT$ edges; how many are required after simplification?
As above, the expected number of edges is bounded by $2N-2$ plus four times the mean marginal tree length.
Roughly, the length of time for which a marginal tree has k tips is
$2N/k(k-1) = 2N(1/(k-1) - 1/k)$,
and so the time to go from N to n tips is $2N(1/(n-1) - 1/(N-1))$.
This implies that after running for time $T$ we expect to
have around $n(T)$ roots, where $n(T) \approx 1 + 2N/(T+2)$.
The total tree length over this time is
$2 N \sum_{k=n(T)+1}^{N-1} 1/k$, which
% is approximately
% \begin{align*}
%     2 N \log\left( \frac{N}{1 + 2 N/(T+2)} \right) .
% \end{align*}
leads to an upper bound on the number of edges of
\begin{align*}
    2 N \left( 1 + 4 \log\left( \frac{NT}{T + 2 N} \right)\right) .
\end{align*}
This holds up quite well in practice --
the number of edges actually required for 50 independent Wright--Fisher simulations,
as a function of time (obtained by running simplify every generation)
is shown in Figure~\ref{fig:num_edges}, compared to this prediction.

@@

In this paper, we have shown that storing pedigrees
and associated recombination events
in a forwards-time simulation
not only results in having available a great deal more information about the simulation,
but also can speed up the simulation by orders of magnitude.
To make this feasible,
we have described how to efficiently store this information in numerical tables,
and have described a fundamental algorithm for simplificication of tree sequences.
Conceptually, recording of genealogical and recombination events
can happen independently of the details of simulation;
for this reason, we provide a well-defined and well-tested API in Python
for use in other code bases (a C API is also planned).

The tree sequences produced by default by this method
are very compact, storing genotype \emph{and} genealogical information
in a small fraction of the space taken by a compressed VCF file.
The format is also highly efficient to process,
leading to advantages for downstream analysis as well.
This is because many algorithms to compute statistics of interest for population genetics
are naturally expressed in terms of tree topologies,
and so can be quickly computed from the trees underlying the tree sequence format.
For example, pairwise nucleotide diversity $\pi$, is the average density of
differences between sequences in the sample.
To compute this directly from sequence data at $m$ sites in $n$ samples
requires computing allele frequencies, taking $O(nm)$ operations.
By using the locations of the mutations on the marginal trees,
and the fact that these are correlated,
sequential tree algorithms similar to those in~\citep{kelleher2016efficient}
can do this in roughly $O(n + m \log n)$ operations.
The \msprime\ API provides a method to compute $\pi$ among arbitrary subsets of the
samples in a tree sequence, which took about 1.2 seconds
applied to the example tree sequence above with 1.1 million mutations
in 200 megabases for 500,000 samples.

@@

The Tables API is primarily designed to facilitate efficient interchange of
data between programs or between different modules of the same program.
\jda{should capitalize Tables, or not, as above}
Following the current best-practises [citations: apache arrow, etc] data is stored
in a columnar format.
The principal advantage of this for our purposes is that it allows for very efficient
interchange of large amounts of numerical data. In principle, this enables
zero-copy semantics, where a data consumer can read the information directly
from the memory of a producer without incurring the overhead of a copy
[citation?] Our implementation uses the numpy C API [citation] to efficiently copy
data from Python into the low-level C library used to manipulate
tree sequences.
Thus, by using a simple numerical
encoding of tree topologies and contiguous arrays of data to store table
columns, we can achieve data transfer rates that would be impossible under
any text-based approach while retaining excellent portability.

@@

\jk{Moved from above}
More concretely,
the algorithm works by moving back through time,
processing each parent in the input tree sequence in chronological order.
The main state of the algorithm at each point in time is a set of ancestral lineages,
and each lineage is a collection of ancestral segments.
An ancestral segment $(\ell, r, u)$ is found in a lineage
if the output node $u$ inherits the genomic interval $[\ell, r)$ from that lineage.
% These segments are stored in a collection of linked lists,
% one list of segments for each ancestral lineage present at that time.
(These lineages are the ``colors'' above.)
We also maintain a map $A$ such that $A_j$ is segment chain for node $j$ in
the input tree sequence.
Crucially, the time required to run the algorithm is linear in the number of edges of the input tree sequence.
